---
Project: "Comprehensive Assessment: Puerto Rico Hurricane Mortality"
Script:  "pr_hurr_scrubber.R"
Purpose: "Script to ingest PDF data and clean it so that is suitable for queries"
Author:  "Larry Mannings"
Date:    "2026-03-06"
---



**Part I --- Preparation ---**

```{r prep, message=FALSE, warning=FALSE}

# 1. Clear and set the environment

rm(list = ls())
options(digits = 3)    # report 3 significant digits

# 2. Load libraries

library(tidyverse)  # Data manipulation
library(pdftools)   # Extract text from a PDF file
library(stringr)    # wrapper function for commmon string operations

# 3. Global variables / paths
INPUT_FILE <- system.file("extdata", "RD-Mortality-Report_2015-18-180531.pdf", package="dslabs")
```

**Part II --- Data Wrangling ---**

This section walks through step by step the process of extracting data from a demographics PDF tracking Puerto Rico mortality data and converting this data into a tidy format.

*1. Find and open the the file directly from RStudio (Windows executable)*
```{r results='hide'}

system("cmd.exe", input = paste("start", INPUT_FILE))
```

Inspection of the file shows that it is a PDF file with tables and graphs. Extraction of the data seems feasible.

*2. Read the PDF file and store the results in an object called `txt`.*

```{r message=FALSE, warning=FALSE}
txt <- pdf_text(INPUT_FILE)
```

Inspection of the object `txt` reveals a character string of length 12 (number of
pages) corresponding to the text on page. The mortality data is in there somewhere.


*3. Extract the ninth page from the object `txt` (txt[9]), and split the string into separate substrings at the occurence of a pattern match (\n)*
```{r message=FALSE, warning=FALSE}
x <- str_split(txt[9], "\\n")   # return a list of substrings
str(x)  # display the structure of the object x
```

It can be observed that the object `x` is a list of size 1 containing a character string of 41 elements. 

*4. Lets look at the contents of the list item.*
```{r message=FALSE, warning=FALSE}
s <- x[[1]] # assign the contents of the first list element to s
str(s)  # display the structure of the object s
```

*5. Now lets print a the first part of the object `s`*

```{r message=FALSE, warning=FALSE}

head(s) #display first 6 elements of the vector
```

When substrings in s we obtained above, we see white space before and after the other characters. These will need to be trimmed before attempting to split the substrings further.

*6. Trim spaces at the start and end of strings in `s` and assign the result to `s` again.*

```{r message=FALSE, warning=FALSE}
s <- str_trim(s)    # remove white space from start and end of all substrings
head(s)
```

Now we want to identify the header row by searching for the index of rows that contain header data. We know the header row contains years, so we'll start there.

*7. Find the row with header data and assign to object `header_index`*
```{r message=FALSE, warning=FALSE}
header_index <- str_which(s, "2015")[1] # find index of first row containing "2015"
header_index
```

We want to extract two objects from the header row; month will store the month and header will store the column names

*8. Save the content of the header row into an object called header, then use string operations to extract the objects we need.*
```{r message=FALSE, warning=FALSE}
header <- s[header_index] # extract the string with header data from s
header <- str_split(header, "\\s+", simplify = TRUE) # split header into substrings separated by one or more spaces
month <- header[1]  # assign first element of header to month
header <- header[-1] # remove first element from the header vector
```

Let's look at the tail data of the s object
```{r message=FALSE, warning=FALSE}
tail(s)
```

Notice that towards the end of the page defined by `s` there is a row containing "Total" followed by rows containing other summary statistics. 

*9. Create an object called tail_index containing the index of the "Total" entry*
```{r message=FALSE, warning=FALSE}
tail_index <- str_which(s, "Total") # find index of row containing "Total"
tail_index
```

Because the PDF page contains graphs with numbers, some of the rows of `s` have just one number (corresponding to the numbers in the y-axis of the plot)

*10. Create an object `n` that contains the count of numbers in each row*

```{r message=FALSE, warning=FALSE}
n <- str_count(s, "\\d+")   # count the number of numbers in string row
n[n == 1]  # return the number of rows with only one number
```

We are now ready to remove entries from the rows that we know we don't need.

*11. Remove the entry `header_index` and everything before it. Also remove entries with only one number and the tail_index and all content that comes after it.*

```{r message=FALSE, warning=FALSE}
num_index <- which(n == 1) # return the indices of the s vector that have only one number
s_trunc <- s[-c(1:header_index, num_index, tail_index:length(s))] #keep all 
# elements in s except for specified indices
length(s_trunc) # return the length of truncated vector
```

We are now ready to remove all text that is not a digit or space. You want to use regular expressions (*regex*) and `str_remove_all()` function here.

*12. Remove all text from `s_trunc` that is not either a number or a space*
```{r message=FALSE, warning=FALSE}
s <- str_remove_all(s_trunc, "[^\\d\\s]")
head(s, 10) # display first 10 lines of s
```

Inspecting s again, we see that data corresponding to useful information from PDF table corresponds to the first (date) and second through fifth (mortality count) columns. Now we want to convert the `s` object into a data matrix containing just the day and mortality count data:

```{r message=FALSE, warning=FALSE}
s <- str_split_fixed(s, "\\s+", n = 6)[,1:5] # convert string vector into data matrix
```

We are almost ready to finish. We need to add the column names for day and mortality back to the matrix. we also want to add a column for the month and convert all vales to numeric.

*13. Add column names `date` and header. Add column for `month`. Convert all values to numeric.*

```{r message=FALSE, warning=FALSE}
colnames(s) <- c("day", header) # set column names for date and mortality data columns
m <- as.matrix(rep(month, times = length(s[,1]))) # instantiate month column
colnames(m) <- ("month")
s <- cbind(m, s) # add month column to s
tab <- as.tibble(s) |> mutate_at(-1, parse_number) # convert all but month data to numeric
```

Our table is now suitable for querying using data manipulation grammar

*13a. Compute the mean number of deaths per day in Sep 2015*
```{r message=FALSE, warning=FALSE}
mean(tab$"2015")
```

*13b. What is the mean number of deaths per day in Sep. 2016*
```{r message=FALSE, warning=FALSE}
mean(tab$"2016")
```

*13c. What was the mean number of deaths per day from September 1-19, 2017, before the hurricane hit?*
```{r message=FALSE, warning=FALSE}
mean(tab$"2017"[1:19])
```

*13d. What was the mean number of deaths per day from September 20-30, 2017, after the hurricane hit?*
```{r message=FALSE, warning=FALSE}
mean(tab$"2017"[20:30])
```

Let's finish up by changing `tab` to a tidy format.

*13. Gather columns into key-value pairs. Specifically, gather all the separate year columns and stack them vertically. Associate the respective mortality count as the Value data.*
```{r message=FALSE, warning=FALSE}
tab2 <- tab %>% gather(year, deaths, -c(day, month)) %>%
    mutate(deaths = as.numeric(deaths))
tab2
```

**Part III  --- Visualizing the data ---**

Lastly, we plot the tidy data to help us answer key questions related to how mortality data adjacent to the date Hurricane Maria made landfall compare with data from unaffected dates.


*14. Make a plot of deaths versus day with color to denote year. Exclude 2018  since we have no data. Add a vertical line at day 20, the day that Hurricane María hit in 2017.*
```{r fig.align = "center", message=FALSE, warning=FALSE}
tab_2015 <- tab2 |> filter(year == 2015) # 2015 data
tab_2016 <- tab2 |> filter(year == 2016) # 2016 data
tab_2017 <- tab2 |> filter(year == 2017) # 2017 data

plot(tab_2015$day, tab_2015$deaths, # Plot Dates vs. # of Deaths (2015)
    col = "blue",
    pch = 16,
    xlab = "Month Date", 
    ylab = "Number of Deaths",
    main = "Comparison of Deaths: 2015 to 2017",
    ylim = c(0, max(tab_2015$deaths, tab_2016$deaths, tab_2017$deaths))) 
points(tab_2016$day, tab_2016$deaths, # overlay 2016 data
       col = "green",
       pch = 16) 
points(tab_2017$day, tab_2017$deaths, # overlay 2017 data
       pch = 16,
       col = "orange")
abline(v = 20, col = "red")  # Insert vertical line to mark Hurr. Maria date
legend("right", legend = c("2015 Data", "2016 Data", "2017 Data"),
       col = c("blue", "green", "orange"),
       pch = 16)
```

From observing the plot above, we can make the following observations:

* September 2015 and 2016 deaths by day are roughly equal to each other.
* After the hurricane in September 2017, there were over 100 deaths per day every day for the rest of the month.
* No days before September 20, 2017 have over 100 deaths per day.

**Part IV  --- Conclusion ---**

With consideration of the empirical evidence pertaining to mortality data adjacent to date Hurricane Maria made landfall, it can be concluded that the hurricane had an _*immediate*_ impact on daily mortality in comparison to unaffected days.